Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's \emph{Irises}

### 2020-09-12 penguins.Rmd
update intro

 The famous *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education. While the **iris** measurements were originally collected by botanist Edgar Anderson, their use in Ronald A. Fisher's eugenics research has created momentum to address the dataset's racist use and/or replace **iris** with alternate data in teaching materials. Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data. Here, we introduce the **palmerpenguins** R package and describe the included **penguins** data. We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g. PCA), cluster analysis and classification (e.g. by k-means).

@@

In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935]. The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning. Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data. As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics and data science course materials and tutorials. **Sounds great. Why consider replacing iris?**

@@

There is recent momentum to address Fisher’s racist legacy in statistics. In June 2020, the Committee of Presidents of Statistical Societies’ R.A. Fisher Award and Lectureship was replaced by the Distinguished Achievement Award and Lectureship, to “advance a more just, equitable, diverse, and inclusive statistical community” [@noauthor_institute_2020]. Cambridge University’s Gonville and Caius College recently announced plans to remove a stained-glass window celebrating Fisher from a campus dining hall [@noauthor_sir_2020]. In the R community, there are calls to (1) address the *Iris* dataset's use in Fisher’s eugenics work, and/or (2) consider using a different dataset [@poisot_timothee_its_2020, @garrick_aden-buie_lets_2020]. This paper describes (2): a suitable alternative dataset for educators wanting to replace the *Iris* data in their teaching materials.

Considering *Iris* data usage in data science and statistics materials, we established the following criteria for a suitable replacement dataset:

- Available by appropriate license
- Features intuitive subjects and variables understandable to learners across fields
- Manageable (but not trivial) in size
- Minimal data cleaning and pre-processing required for most analyses
- Real-world (not manufactured) data
- Similar opportunities for teaching and learning R, data science, and statistical skills

Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: an approachable and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER). By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin measurements can replace Anderson's *Iris* data for many use cases in statistics and data science education.

@@

## Other alternatives to iris

## Discussion

Educators have three options regarding **iris** use in course materials:

1. Use **iris** without addressing its use in Fisher's eugenics research
2. Use **iris** *and* address its use in Fisher's eugenics research
3. Replace **iris** with an alternative dataset offering similar teaching and learning opportunities

Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the **palmerpenguins** R package, can replace **iris** for a number of common use cases in data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means. In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.

The **iris** dataset is widespread today because it was employed by Fisher to advance statistical methods for eugenics research. While the *Irises* were described in Anderson's 1935 article in *The Bulletin of the American Iris Society* [@anderson_irises_1935], the measurements were published in full in Fisher's 1936 *Annals of Eugenics* publication [@fisher_use_1936] - a journal for which Fisher served as editor before becoming chair of the Department of Eugenics at University College London.

@@

>>>>

 Edgar Anderson's famous 1935 *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education. Due to a lack of metadata, non-intuitive variables (e.g. Sepal.Width), its publication in (and inextricable connection with) Fisher's 1936 paper in *Annals of Eugenics*, and perfect structure (e.g. no missing values) unrepresentative of most real-world environmental data, a modern alternative to **iris** is needed. Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data. Here, we introduce the **palmerpenguins** R package and describe the included **penguins** data. We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g. PCA), cluster analysis and classification (e.g. by k-means).

@@

In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935]. The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning. Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data. As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, and data science courses and materials.

There are several reasons that practitioners and educators may want to retire Anderson's *Iris* dataset in modern data science materials, including:

- Anderson's  *Iris* dataset entirely lacks metadata, limiting meaningful interpretation and discussion of research methods, analyses, and outcomes
- Two of its five variables, `Sepal.Width` and `Sepal.Length`, are not intuitive to most users
- In 1936, the *Iris* data were published by eugenicist R.A. Fisher in *Annals of Eugenics*, and are now inextricably linked to Fisher's eugenics work; in fact, the dataset is commonly referred to as *Fisher's Iris data*
- The *Iris* dataset contains only complete cases (i.e. no missing values) and equal sample sizes (*n* = 50 for each *Iris* species) - more perfect than real-world datasets learners are likely to encounter "in the wild"
- The single categorical variable (`Species`) limits options for analyses

<!---**Sounds great. Why consider replacing iris?**

@@

There is recent momentum to address Fisher’s racist legacy in statistics. In June 2020, the Committee of Presidents of Statistical Societies’ R.A. Fisher Award and Lectureship was replaced by the Distinguished Achievement Award and Lectureship, to “advance a more just, equitable, diverse, and inclusive statistical community” [@noauthor_institute_2020]. Cambridge University’s Gonville and Caius College recently announced plans to remove a stained-glass window celebrating Fisher from a campus dining hall [@noauthor_sir_2020]. In the R community, there are calls to (1) address the *Iris* dataset's use in Fisher’s eugenics work, and/or (2) consider using a different dataset [@poisot_timothee_its_2020, @garrick_aden-buie_lets_2020]. --->

Given these motivations to retire *Iris*, we sought to identify, then make readily available, a suitable alternative dataset. Considering *Iris* usage in data science and statistics materials, we established the following criteria for a suitable alternative:

- Available by appropriate license (ideally, CC0 "no rights reserved")
- Features intuitive subjects and variables that are interesting and understandable to learners across fields
- Complete metadata and documentation
- Manageable (but not trivial) in size
- Minimal data cleaning and pre-processing required for most analyses
- Real-world (not manufactured), modern data
- Similar opportunities for teaching and learning R, data science, and statistical skills
- Can easily replace *Iris* for most use cases

Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: a refreshing, approachable and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER). By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin measurements are an ideal substitute for Anderson's *Iris* data for many use cases in statistics and data science education.

@@

@@

Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the **palmerpenguins** R package, offer a near drop-in replacement for **iris** in a number of common use cases for data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means. In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.

### 2020-09-12 penguins.tex
update intro

\abstract{%
The famous \emph{Iris} dataset, which exists in the base R
\textbf{datasets} package as \textbf{iris}, is ubiquitous in statistics
and data science education. While the \textbf{iris} measurements were
originally collected by botanist Edgar Anderson, their use in Ronald A.
Fisher's eugenics research has created momentum to address the dataset's
racist use and/or replace \textbf{iris} with alternate data in teaching
materials. Penguin morphological measurements for three species of
\emph{Pygoscelis} penguins that breed on islands throughout the Palmer
Archipelago, Antarctica, provide an approachable, charismatic and near
drop-in replacement for the \textbf{iris} data. Here, we introduce the
\textbf{palmerpenguins} R package and describe the included
\textbf{penguins} data. We directly compare the two datasets for
selected analyses to demonstrate that R users, in particular teachers
and learners currently using the \textbf{iris} data, can switch to the
Palmer Archipelago penguins for many use cases including data wrangling
and visualization, correlation, regression, hypothesis testing,
multivariate analysis (e.g.~PCA), cluster analysis and classification
(e.g.~by k-means).
}

@@

In 1935, American botanist Edgar Anderson measured petal and sepal
structural dimensions (length and width) for 50 flowers from three Iris
species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris
virginica} \citep{anderson_irises_1935}. The manageable but non-trivial
size (5 variables and 150 total observations) and characteristics of the
data make it amenable for introducing a wide range of statistical
methods including data wrangling, visualization, regression,
multivariate exploration, and machine learning. Anderson's \emph{Iris}
dataset is built into a number of software packages including the
auto-installed datasets package in R \citep{r_core_team_r_2019},
Python's scikit-learn machine learning library
\citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS
Institute, Cary NC), which have facilitated widespread use of the data.
As a result, eighty-five years after the data were initially published,
Anderson's \emph{Iris} measurements are ubiquitous in statistics and
data science course materials and tutorials. \textbf{Sounds great. Why
consider replacing iris?}

In 1936, one year after being described in the Bulletin of the American
Iris Society, Anderson's \emph{Iris} measurements were published in full
by eugenicist and statistician Ronald A. Fisher in \emph{Annals of
Eugenics} \citep{fisher_use_1936}. Regardless of Anderson's initial
motivation, the data today remain inextricably linked to Fisher's
eugenics research and are even commonly, if unfairly, referred to as
``Fisher's iris data.'' For example, an August 2020 search for ``Iris
flower data set'' in Wikipedia returns an article beginning with ``The
iris flower data set or Fisher's iris data\ldots{}'', then later in the
article: ``It is sometimes called Anderson's Iris data set''
\citep{wikipedia_iris_2020}. The \emph{Iris} dataset is similarly
credited to Fisher in statistical computing literature
\citep[\citet{wang_matlab_2015}, \citet{woods_how_2015},
\citet{chen_unsupervised_2018}]{trendafilov_simple_2009}.

There is recent momentum to address Fisher's racist legacy in
statistics. In June 2020, the Committee of Presidents of Statistical
Societies' R.A. Fisher Award and Lectureship was replaced by the
Distinguished Achievement Award and Lectureship, to ``advance a more
just, equitable, diverse, and inclusive statistical community''
\citep{noauthor_institute_2020}. Cambridge University's Gonville and
Caius College recently announced plans to remove a stained-glass window
celebrating Fisher from a campus dining hall \citep{noauthor_sir_2020}.
In the R community, there are calls to (1) address the \emph{Iris}
dataset's use in Fisher's eugenics work, and/or (2) consider using a
different dataset
\citep[\citet{garrick_aden-buie_lets_2020}]{poisot_timothee_its_2020}.
This paper describes (2): a suitable alternative dataset for educators
wanting to replace the \emph{Iris} data in their teaching materials.

Considering \emph{Iris} data usage in data science and statistics
materials, we established the following criteria for a suitable
replacement dataset:

\begin{itemize}
\tightlist
\item
  Available by appropriate license
\item
  Features intuitive subjects and variables understandable to learners
  across fields
\item
  Manageable (but not trivial) in size
\item
  Minimal data cleaning and pre-processing required for most analyses
\item
  Real-world (not manufactured) data
\item
  Similar opportunities for teaching and learning R, data science, and
  statistical skills
\end{itemize}

Here, we describe an alternative to Anderson's \emph{Iris} data that
largely satisfies these criteria: an approachable and charismatic
dataset containing real-world morphological data for three
\emph{Pygoscelis} penguin species that breed throughout the Western
Antarctic Peninsula region, made available through the Long-Term
Ecological Research Network (US LTER). By comparing data structure,
size, and a range of analyses side-by-side for the two datasets, we
demonstrate that the Palmer Archipelago penguin measurements can replace
Anderson's \emph{Iris} data for many use cases in statistics and data
science education.

@@

\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
\centering
\begin{tabular}[t]{c|c|c|c|c|c|c|c}
\hline
\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
\cline{1-4} \cline{5-8}
Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
\hline
1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
\hline
2 & 4 & 54 & 6 & 2 & NA & 2 & 46\\
\hline
3 & 147 & 5 & 1 & 3 & 50 & NA & NA\\
\hline
\end{tabular}
\end{table}

@@

\hypertarget{other-alternatives-to-iris}{%
\subsection{Other alternatives to
iris}\label{other-alternatives-to-iris}}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

Educators have three options regarding \textbf{iris} use in course
materials:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Use \textbf{iris} without addressing its use in Fisher's eugenics
  research
\item
  Use \textbf{iris} \emph{and} address its use in Fisher's eugenics
  research
\item
  Replace \textbf{iris} with an alternative dataset offering similar
  teaching and learning opportunities
\end{enumerate}

Here, we have shown that structural size measurements for Palmer
Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
in the \textbf{palmerpenguins} R package, can replace \textbf{iris} for
a number of common use cases in data science and statistics education
including exploratory data visualization, linear correlation and
regression, PCA, and clustering by k-means. In addition, teaching and
learning opportunities in \textbf{penguins} are increased due to a
greater number of variables, missing values, unequal sample sizes, and
Simpson's paradox examples.

The \textbf{iris} dataset is widespread today because it was employed by
Fisher to advance statistical methods for eugenics research. While the
\emph{Irises} were described in Anderson's 1935 article in \emph{The
Bulletin of the American Iris Society} \citep{anderson_irises_1935}, the
measurements were published in full in Fisher's 1936 \emph{Annals of
Eugenics} publication \citep{fisher_use_1936} - a journal for which
Fisher served as editor before becoming chair of the Department of
Eugenics at University College London.

@@

>>>>

\abstract{%
Edgar Anderson's famous 1935 \emph{Iris} dataset, which exists in the
base R \textbf{datasets} package as \textbf{iris}, is ubiquitous in
statistics and data science education. Due to a lack of metadata,
non-intuitive variables (e.g.~Sepal.Width), its publication in (and
inextricable connection with) Fisher's 1936 paper in \emph{Annals of
Eugenics}, and perfect structure (e.g.~no missing values)
unrepresentative of most real-world environmental data, a modern
alternative to \textbf{iris} is needed. Penguin morphological
measurements for three species of \emph{Pygoscelis} penguins that breed
on islands throughout the Palmer Archipelago, Antarctica, provide an
approachable, charismatic and near drop-in replacement for the
\textbf{iris} data. Here, we introduce the \textbf{palmerpenguins} R
package and describe the included \textbf{penguins} data. We directly
compare the two datasets for selected analyses to demonstrate that R
users, in particular teachers and learners currently using the
\textbf{iris} data, can switch to the Palmer Archipelago penguins for
many use cases including data wrangling and visualization, correlation,
regression, hypothesis testing, multivariate analysis (e.g.~PCA),
cluster analysis and classification (e.g.~by k-means).
}

@@

In 1935, American botanist Edgar Anderson measured petal and sepal
structural dimensions (length and width) for 50 flowers from three Iris
species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris
virginica} \citep{anderson_irises_1935}. The manageable but non-trivial
size (5 variables and 150 total observations) and characteristics of the
data make it amenable for introducing a wide range of statistical
methods including data wrangling, visualization, regression,
multivariate exploration, and machine learning. Anderson's \emph{Iris}
dataset is built into a number of software packages including the
auto-installed datasets package in R \citep{r_core_team_r_2019},
Python's scikit-learn machine learning library
\citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS
Institute, Cary NC), which have facilitated widespread use of the data.
As a result, eighty-five years after the data were initially published,
Anderson's \emph{Iris} measurements are ubiquitous in statistics,
computational methods, and data science courses and materials.

There are several reasons that practitioners and educators may want to
retire Anderson's \emph{Iris} dataset in modern data science materials,
including:

\begin{itemize}
\tightlist
\item
  Anderson's \emph{Iris} dataset entirely lacks metadata, limiting
  meaningful interpretation and discussion of research methods,
  analyses, and outcomes
\item
  Two of its five variables, \texttt{Sepal.Width} and
  \texttt{Sepal.Length}, are not intuitive to most users
\item
  In 1936, the \emph{Iris} data were published by eugenicist R.A. Fisher
  in \emph{Annals of Eugenics}, and are now inextricably linked to
  Fisher's eugenics work; in fact, the dataset is commonly referred to
  as \emph{Fisher's Iris data}
\item
  The \emph{Iris} dataset contains only complete cases (i.e.~no missing
  values) and equal sample sizes (\emph{n} = 50 for each \emph{Iris}
  species) - more perfect than real-world datasets learners are likely
  to encounter ``in the wild''
\item
  The single categorical variable (\texttt{Species}) limits options for
  analyses
\end{itemize}

Given these motivations to retire \emph{Iris}, we sought to identify,
then make readily available, a suitable alternative dataset. Considering
\emph{Iris} usage in data science and statistics materials, we
established the following criteria for a suitable alternative:

\begin{itemize}
\tightlist
\item
  Available by appropriate license (ideally, CC0 ``no rights reserved'')
\item
  Features intuitive subjects and variables that are interesting and
  understandable to learners across fields
\item
  Complete metadata and documentation
\item
  Manageable (but not trivial) in size
\item
  Minimal data cleaning and pre-processing required for most analyses
\item
  Real-world (not manufactured), modern data
\item
  Similar opportunities for teaching and learning R, data science, and
  statistical skills
\item
  Can easily replace \emph{Iris} for most use cases
\end{itemize}

Here, we describe an alternative to Anderson's \emph{Iris} data that
largely satisfies these criteria: a refreshing, approachable and
charismatic dataset containing real-world morphological data for three
\emph{Pygoscelis} penguin species that breed throughout the Western
Antarctic Peninsula region, made available through the Long-Term
Ecological Research Network (US LTER). By comparing data structure,
size, and a range of analyses side-by-side for the two datasets, we
demonstrate that the Palmer Archipelago penguin measurements are an
ideal substitute for Anderson's \emph{Iris} data for many use cases in
statistics and data science education.

@@

\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
\centering
\begin{tabular}[t]{c|c|c|c|c|c|c|c}
\hline
\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
\cline{1-4} \cline{5-8}
Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
\hline
1 & 4 & 54 & 6 & 1 & 50 & NA & NA\\
\hline
2 & 147 & 5 & 1 & 2 & NA & 48 & 4\\
\hline
3 & NA & 9 & 116 & 3 & NA & 2 & 46\\
\hline
\end{tabular}
\end{table}

@@

@@

Here, we have shown that structural size measurements for Palmer
Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
in the \textbf{palmerpenguins} R package, offer a near drop-in
replacement for \textbf{iris} in a number of common use cases for data
science and statistics education including exploratory data
visualization, linear correlation and regression, PCA, and clustering by
k-means. In addition, teaching and learning opportunities in
\textbf{penguins} are increased due to a greater number of variables,
missing values, unequal sample sizes, and Simpson's paradox examples.
